---
title: "Spatial Data"
author: "Anastasia Kunz"
date: "2/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(sf)
library(tmap)


```

## Spatial Data Wrangling and Visualization

#### 1. California Counties 

```{r}
ca_counties <- (read_sf(here("data", "ca_counties", "CA_Counties_TIGER2016.shp")))
```

make a subset with only county_name and land_area

```{r}
ca_subset <- ca_counties %>% 
  select(NAME, ALAND) %>% 
  rename(county_name = NAME, land_area = ALAND)
```

geometry data is "sticky" even though we didnt select it, it sticks around

### Set the CRS
check existing crs for the spatial data

```{r}
ca_subset %>%  st_crs()
```

#### Look at the data
plot using geom_sf() function, can update aesthetics like ggplot. 

Update color based on land area

```{r}
ggplot(data = ca_subset) +
  geom_sf(aes(fill = land_area), color = "white", siize = 0.1) +
  theme_void()+
  scale_fill_gradientn(colors = c("cyan", "blue", "purple"))
```

#### 2. Invasive red sesbania records
invasive plant in california

```{r}
sesbania <- read_sf(here("data", "red_sesbania", "ds80.shp"))

sesbania %>% st_crs()
```

because they dont match we will want to reproject it to match using st_transform()

```{r}
sesbania <- st_transform(sesbania, 3857)

sesbania %>% st_crs()
```

Same projection!!!

#### 3. Plot them both together!
note i actually had to put "data = xxx" in my geom... dont usually have to do that

```{r}
ggplot()+
  geom_sf(data = ca_subset)+
  geom_sf(data = sesbania, size = 1, color = "red")
```


####  Data Wrangling

**Spatial Joins**

st_join will join the two datasets

```{r}
ca_sesbania <- ca_subset %>% 
  st_join(sesbania)
```

Now we can find the counts of sesbania in each county

```{r}
sesbania_counts <- ca_sesbania %>% 
  count(county_name)
```

now we can create a chloropleth map using the number of records for sesbania as fill color (instead of land area from previous graph)

```{r}
ggplot(data = sesbania_counts) +
  geom_sf(aes(fill = n), color = "white", size = 0.1)+
  scale_fill_gradientn(colors = c( "lightgray", "orange", "red")) +
  theme_minimal()+
  labs(fill = "Number of s. punicea records")
```

Focus on a county do some wrangling. Look at solano county with lots of sesbania

```{r}
solano_sesbania <- sesbania %>% 
  filter(COUNTY == "Solano")

solano <- ca_subset %>% 
  filter(county_name == "Solano")

ggplot()+
  geom_sf(data = solano)+
  geom_sf(data = solano_sesbania)

```

#### Make an interactive map with tmap

make one for CA counties (fill aesthetic by land area with the red sesbania locations on the top)

```{r}
#set mode to interactive
tmap_mode(mode = "view")

# Make map with polygon fill color updated by the var land_area, updating the color pallette to "Bu Gn" then add another shape layer for the sesbania records as dots

tm_shape(ca_subset)+
  tm_fill("land_area", palette = "BuGn") +
  tm_shape(sesbania)+
  tm_dots()+
  tm_basemap("Stamen.Watercolor")



```

